Efficient saddle point search - Binary image transistion state search

### 2022-02-01 manuscript.tex
Made changes to match PRL guidelines

\documentclass[aps,twocolumn]{revtex4-1}

@@

\begin{document}
\title{Efficient saddle point search - Binary image transistion state search}
\author{Sam Avis}
\author{Jack Panter}
\author{Halim Kusumaatmaja}

@@

\begin{figure}[htb]
  \centering
  \includegraphics{fig/toy2d.pdf}
  \caption{
    (a) The trajectories of the two states under the BITSS method for a simple 2D potential with two minima (blue dots).
        The transition state is shown by an red dot, and the minimum energy pathway by the dashed line.
    (b) A snapshot of the BITSS minimisation showing the driving forces on each state due to the energy constraint, $\bm{F}_E$, and distance constraint, $\bm{F}_D$, with $E_1 < E_2$ and $d(\bm{x}_1,\bm{x}_2) > d_i$.
    (c) The final configuration of the BITSS method showing the two states in orange, the transition state in red, and the negative curvature eigenvector, $\bm{\hat{\tau}}$.
  }
  \label{fig:toy2d}
\end{figure}

@@

\topic The method is sensitive to the choice of the constraint coefficients, $\kappa_d$ and $\kappa_e$, with optimal values depending upon the specific problem.
Also, as the separation between the states decreases the coefficients will have to change, making them difficult to assign manually.
Consequently, we automate their assignment using information from the system.
They are initially calculated at the start of each minimisation, and regularly recalculated throughout (once per 100 iterations are used in this work).
The coefficients are chosen such that each term in the gradient of \cref{eq:bitss} is of a similar size, providing the following expressions,
\begin{gather}
  \kappa_e = \frac {\alpha} {2 \Delta E},
  \label{eq:ke}
  \\
  \kappa_d = \text{max} \left(
    \frac {\sqrt{|\bm{\nabla} E_1|^2 + |\bm{\nabla} E_2|^2}} {2\sqrt{2} \beta d_i} \; , \;
    \frac{\Delta E}{\beta d_i^2} \right),
  \label{eq:kd}
\end{gather}
where $\alpha$ and $\beta$ are parameters with recommended values of $\alpha = 10$ and $\beta = 0.1$ (see supplementary information).
$\Delta E$ is an estimation for the current energy barrier which is evaluated by linearly interpolating between the two current states.
For systems where the interpolation deviates significantly from the MEP this estimate can be much too large.
In which case, a maximum estimated value for the barrier can be provided, which then decreases proportionally with $d_i$.

@@

% This is here to prevent it being put on the wrong page
\begin{figure*}[tb]
  \centering
  \includegraphics{fig/speedtest-a.pdf}%
  \includegraphics{fig/speedtest-b.pdf}%
  \includegraphics{fig/speedtest-c.pdf}%
  \caption{
    The rate of convergence to the transition state for (a) a Lennard-Jones seven-particle cluster, (b) cylindrical shell buckling, (c) wetting of a chemically-striped surface.
    Above are the two minimum energy states and the transition state between them.
    Below is shown the convergence to the transition state as a function of the number of gradient calculations using the BITSS (black line) and string methods.
    The string method is repeated with a differing number of images along the string, the number of which is listed in the legend in (c).
  }
  \label{fig:speedtest}
\end{figure*}

\topic The typical trajectory of the two states in the BITSS method is shown in \cref{fig:toy2d}a.
Initially, the lower energy state jumps up to satisfy the equal energy constraint and then moves to reduce the separation as much as possible without increasing it's energy.
Then, the two states begin to converge directly towards one another, before being deflected in the direction of a saddle in the potential.
Consequently, if there are multiple possible pathways between two states, BITSS will be biased towards identifying pathways that are more direct or with lower energy.
As the two states approach the transition state, the path they take begins to closely follow the minimum energy pathway (MEP), which is the path that follows the gradient from the minima and passes through the transition state.
Therefore, the final two states are positioned in the direction of the negative curvature eigenvector, $\bm{\hat{\tau}}$, as shown in \cref{fig:toy2d}c.
So it is also possible to obtain the eigenvector and eigenvalue using BITSS.

\topic Once BITSS has identified the transition state it is possible to find the full MEP.
This is done by tracing the trajectory of downhill minimisations from either side of the transition state, which are provided by the two final states of BITSS.
To ensure this pathway is sufficiently smooth the maximum step size of the minimisation should be restricted.

@@

\topic The final system involves a droplet situated on a chemically striped surface with both hydrophilic and hydrophobic regions.
This has two possible states, with the droplet either upon one hydrophilic stripe, or straddling across two.
This is modelled on a 200x200 2D grid using a phase-field model \cite{Panter2019b}, which has an order parameter, $\phi(\bm{r})$, representing the phase of the liquid ($\phi=-1$ for gas, $\phi=1$ for liquid).
The energy functional has four separate terms,
\begin{equation}
  E[\phi] = E_B + E_I + E_S + E_V.
\end{equation}
The first term, $E_B$, uses a double well potential to set values of $\phi=\pm1$ in the bulk.
The second term then provides the interfacial energy between the liquid and gas by imposing an energy penalty to changes in $\phi$.
$E_S$ is the solid-liquid interaction energy, which sets the contact angles of the hydrophilic and hydrophobic regions to 60\si{\degree} and 110\si{\degree}, respectively.
Finally, $E_V$ constrains the volume of the liquid drop by penalising any variation from 11\% of the total volume of the system.

@@

\begin{figure}[tb]
  \centering
  \begin{tabular}[b]{c}
    \includegraphics{fig/changingmesh.pdf}\\
    \includegraphics{fig/differentgrids.pdf}%
  \end{tabular}
  \caption{
    (a) Snapshots of the BITSS method applied to the buckling of a cylinder with a changing mesh.
        The radial displacement relative the unbuckled cylinder is shown, as well as the underlying triangular mesh.
    (b) BITSS applied to the striped wetting example with different resolutions for the two states.
        Each grid cell denotes 50x50 lattice nodes.
        The zoomed axes shows the difference in the fluid interface between the two final states, as well as the approximated transition state (solid black line).
        This is compared to the true transition state found using a high resolution for both states (dashed line).
  }
  \label{fig:adaptivemesh}
\end{figure}

@@

\topic In \cref{fig:flatdiscontinuous}a we demonstrate the BITSS method applied to a 2D potential with regions containing such zero modes (i \& ii).
We see it is able to successfully converge past these flat regions, even with one very close to the transition state (ii).
In these regions there are no driving forces due to the potential and the energy constraint, which use purely local information about the gradient.
However, the distance constraint continues to pull the states together, preventing them from getting stuck.
When only a single state has a zero-gradient mode then the other is likely to slide down the potential slightly, causing a kink in the pathway (iii).
Although, the two states still remain either side of the dividing ridge and so the result will be unaffected.
Indeed, it may only cause BITSS to fail if the zero-gradient mode is positioned at the transition state in the direction of the pathway.
But in this case there is no single transition state, instead the MEP would have a flat section at the top of its energy profile.

\begin{figure}[tb]
  \centering
  \begin{tabular}[b]{c}
    \includegraphics{fig/zeroeigen.pdf}\\
    \includegraphics{fig/discontinuous.pdf}
  \end{tabular}
  \caption{
    (a) Energy profile of the BITSS pathway on a 2D potential with flat regions.
        The minima and transition state are marked by blue and red dots respectively.
        Points of interest are labelled by i-iii.
        Top inset: A zoomed in view around the transition state.
        Bottom inset: The pathway taken, with the edges of the flat regions marked by dashed lines.
    (b) Disconnectivity graph of the energy landscape for a seven-particle cluster with a discontinuous hard-core pair potential.
    (c) Plot of the discontinuous potential.
        The repulsive section of the usual Lennard-Jones potential is also shown in grey.
  }
  \label{fig:flatdiscontinuous}
\end{figure}

@@

\topic However, BITSS can be used for these systems with undefined gradients with just a couple of alterations.
First, the term using gradients in \cref{eq:kd} can be ignored and $\kappa_d$ can instead be calculated using just the second term.
Next we must use a minimiser that does not require a differentiable optimisation function; for this we use simulated annealing \cite{Kirkpatrick1983}.
This has a chance of randomly jumping one state over the dividing barrier, but we can reduce this probability by setting the initial temperature to $T_0 = \frac{\Delta E}{10}$, and the maximum size of random displacements to $d_\text{max} = d(\bm{x}_1, \bm{x}_2) / 100$.

@@

\bibliographystyle{plain}
\bibliography{../../library.bib}

>>>>

\documentclass[aps,prl,twocolumn,groupedaddress]{revtex4}

@@

\begin{document}
\title{Efficient saddle point search - Binary image transistion state search}
\author{Sam Avis}
\author{Jack Panter}
\author{Halim Kusumaatmaja}
\affiliation{Department of Physics, Durham University, South Road, Durham DH1 3LE, UK}

@@

\begin{figure}[htb]
  \includegraphics{fig/toy2d.pdf}
  \caption{
    (a) The trajectories of the two states under the BITSS method for a simple 2D potential with two minima (blue dots).
        The transition state is shown by an red dot, and the minimum energy pathway by the dashed line.
    (b) A snapshot of the BITSS minimisation showing the driving forces on each state due to the energy constraint, $\bm{F}_E$, and distance constraint, $\bm{F}_D$, with $E_1 < E_2$ and $d(\bm{x}_1,\bm{x}_2) > d_i$.
    (c) The final configuration of the BITSS method showing the two states in orange, the transition state in red, and the negative curvature eigenvector, $\bm{\hat{\tau}}$.
  }
  \label{fig:toy2d}
\end{figure}

@@

\topic The method is sensitive to the choice of the constraint coefficients, $\kappa_d$ and $\kappa_e$, with optimal values depending upon the specific problem.
Also, as the separation between the states decreases the coefficients will have to change, making them difficult to assign manually.
Consequently, we automate their assignment using information from the system.
They are initially calculated at the start of each minimisation, and regularly recalculated throughout (once per 100 iterations are used in this work).
The coefficients are chosen such that each term in the gradient of \cref{eq:bitss} is of a similar size, providing the following expressions,
\begin{gather}
  \kappa_e = \frac {\alpha} {2 \Delta E},
  \label{eq:ke}
  \\
  \kappa_d = \text{max} \left(
    \frac {\sqrt{|\bm{\nabla} E_1|^2 + |\bm{\nabla} E_2|^2}} {2\sqrt{2} \beta d_i} \; , \;
    \frac{\Delta E}{\beta d_i^2} \right),
  \label{eq:kd}
\end{gather}
where $\alpha$ and $\beta$ are parameters with recommended values of $\alpha = 10$ and $\beta = 0.1$ (see supplementary information).
$\Delta E$ is an estimation for the current energy barrier which is evaluated by linearly interpolating between the two current states.
For systems where the interpolation deviates significantly from the minimum energy pathway this estimate can be much too large.
In which case, a maximum estimated value for the barrier can be provided, which then decreases proportionally with $d_i$.

@@

% This is here to prevent it being put on the wrong page
\begin{figure*}[tb]
  \includegraphics{fig/speedtest-a.pdf}%
  \includegraphics{fig/speedtest-b.pdf}%
  \includegraphics{fig/speedtest-c.pdf}%
  \caption{
    The rate of convergence to the transition state for (a) a Lennard-Jones seven-particle cluster, (b) cylindrical shell buckling, (c) wetting of a chemically-striped surface.
    Above are the two minimum energy states and the transition state between them.
    Below is shown the convergence to the transition state as a function of the number of gradient calculations using the BITSS (black line) and string methods.
    The string method is repeated with a differing number of images along the string, the number of which is listed in the legend in (c).
  }
  \label{fig:speedtest}
\end{figure*}

\topic The typical trajectory of the two states in the BITSS method is shown in \cref{fig:toy2d}a.
Initially, the lower energy state jumps up to satisfy the equal energy constraint and then moves to reduce the separation as much as possible without increasing it's energy.
Then, the two states begin to converge directly towards one another, before being deflected in the direction of a saddle in the potential.
Consequently, if there are multiple possible pathways between two states, BITSS will be biased towards identifying pathways that are more direct or with lower energy.
As the two states approach the transition state, the path they take begins to closely follow the minimum energy pathway, which is the path that follows the gradient from the minima and passes through the transition state.
Therefore, the final two states are positioned in the direction of the negative curvature eigenvector, $\bm{\hat{\tau}}$, as shown in \cref{fig:toy2d}c.
So it is also possible to obtain the eigenvector and eigenvalue using BITSS.

\topic Once BITSS has identified the transition state it is possible to find the full minimum energy pathway.
This is done by tracing the trajectory of downhill minimisations from either side of the transition state, which are provided by the two final states of BITSS.
To ensure this pathway is sufficiently smooth the maximum step size of the minimisation should be restricted.

@@

\topic The final system involves a droplet situated on a chemically striped surface with both hydrophilic and hydrophobic regions.
This has two possible states, with the droplet either upon one hydrophilic stripe, or straddling across two.
This is modelled on a 200x200 2D grid using a phase-field model \cite{Panter2019b}, which has an order parameter, $\phi(\bm{r})$, representing the phase of the liquid ($\phi=-1$ for gas, $\phi=1$ for liquid).
The energy functional has four separate terms,
\begin{equation}
  E[\phi] = E_B + E_I + E_S + E_V.
\end{equation}
The first term, $E_B$, uses a double well potential to set values of $\phi=\pm1$ in the bulk.
The second term then provides the interfacial energy between the liquid and gas by imposing an energy penalty to gradients in $\phi$.
$E_S$ is the solid-liquid interaction energy, which sets the contact angles of the hydrophilic and hydrophobic regions to 60\si{\degree} and 110\si{\degree}, respectively.
Finally, $E_V$ constrains the volume of the liquid drop by penalising any variation from 11\% of the total volume of the system.

@@

\begin{figure}[tb]
  \begin{tabular}[b]{c}
    \includegraphics{fig/changingmesh.pdf}\\
    \includegraphics{fig/differentgrids.pdf}%
  \end{tabular}
  \caption{
    (a) Snapshots of the BITSS method applied to the buckling of a cylinder with a changing mesh.
        The radial displacement relative the unbuckled cylinder is shown, as well as the underlying triangular mesh.
    (b) BITSS applied to the striped wetting example with different resolutions for the two states.
        Each grid cell denotes 50x50 lattice nodes.
        The zoomed axes shows the difference in the fluid interface between the two final states, as well as the approximated transition state (solid black line).
        This is compared to the true transition state found using a high resolution for both states (dashed line).
  }
  \label{fig:adaptivemesh}
\end{figure}

@@

\topic In \cref{fig:flatdiscontinuous}a we demonstrate the BITSS method applied to a 2D potential with regions containing such zero modes (i \& ii).
We see it is able to successfully converge past these flat regions, even with one very close to the transition state (ii).
In these regions there are no driving forces due to the potential and the energy constraint, which use purely local information about the gradient.
However, the distance constraint continues to pull the states together, preventing them from getting stuck.
When only a single state has a zero-gradient mode then the other is likely to slide down the potential slightly, causing a kink in the pathway (iii).
Although, the two states still remain either side of the dividing ridge and so the result will be unaffected.
Indeed, it may only cause BITSS to fail if the zero-gradient mode is positioned at the transition state in the direction of the pathway.
But in this case there is no single transition state, instead the minimum energy pathway would have a flat section at the top of its energy profile.

\begin{figure}[tb]
  \begin{tabular}[b]{c}
    \includegraphics{fig/zeroeigen.pdf}\\
    \includegraphics{fig/discontinuous.pdf}
  \end{tabular}
  \caption{
    (a) Energy profile of the BITSS pathway on a 2D potential with flat regions.
        The minima and transition state are marked by blue and red dots respectively.
        Points of interest are labelled by i-iii.
        Top inset: A zoomed in view around the transition state.
        Bottom inset: The pathway taken, with the edges of the flat regions marked by dashed lines.
    (b) Disconnectivity graph of the energy landscape for a seven-particle cluster with a discontinuous hard-core pair potential.
    (c) Plot of the discontinuous potential.
        The repulsive section of the usual Lennard-Jones potential is also shown in grey.
  }
  \label{fig:flatdiscontinuous}
\end{figure}

@@

\topic However, BITSS can be used for these systems with undefined gradients with just a couple of alterations.
First, the term using gradients in \cref{eq:kd} can be ignored and $\kappa_d$ can instead be calculated using just the second term.
Next we must use a minimiser that does not require a differentiable optimisation function; for this we use simulated annealing \cite{Kirkpatrick1983}.
This has a chance of randomly jumping one state over the dividing barrier, but we can reduce this probability by setting the initial temperature to $T_0 = \Delta E / 10$, and the maximum size of random displacements to $d_\text{max} = d(\bm{x}_1, \bm{x}_2) / 100$.

@@

% \bibliographystyle{apsrev4-2}
\bibliography{../../library.bib}
